Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2015-04-05 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Added text to Ch 4  Broke Bibtek

\Title{Repeatability of Pre-Flashover Fire Patterns on Gypsum Wall Board}
\Author{Daniel Madrzykowski}
\Year{2015}

@@

A free standing steel frame wall with a 12.7~mm thick, 2.44~m high by 1.22~m wide, calcium silicate board face was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux. Marinite Type I is the commercial name of the non-combustible board used in these experiments.  It is normally used as a fire-resistant thermal insulation that also has structural strength. 90~\% of the board is composed of calcium silicate and calcium meta-silicate.  The remaining 10~\% of the board is composed of a combination of organic fibers, fiber glass filament and crystalline silica (quartz) ref BNZ MSDS.  Material properties that impact heat transfer were selected from the manufacturer's literature and are presented in the table below ref BNZ Marinite I.

\begin{table}
	\centering
\begin{tabular}{|c|c|c|c|c|}
	\hline Density & \multicolumn{2}{c|}{Thermal Conductivity} & \multicolumn{2}{c|}{Specific Heat}   \\
    \hline (kg/m$^3$) & Temperature ($^\circ$C) & (W/mK)  & Temperature ($^\circ$C)  & (kJ/kg $^\circ$C) \\
	\hline 737  & 24 & 0.13 & 93 & 0.28 \\
	\hline  & 205 & 0.12  & 205 & 0.30 \\
	\hline  & 316 & 0.11 & 316 & 0.32 \\
	\hline  & 425 & 0.12 & 425 & 0.34 \\
	\hline  & 538 & 0.12 &  &  \\
	\hline
\end{tabular}
\caption[Thermal properties for Marinite I]{Thermal properties for Marinite I, as reported by the manufacturer BNZ Marinite I.}
\label{tab:Marinite_Thermal_Properties}
\end{table}

@@

Figures~\ref{TC_Plume_TWNG_comp} through \ref{TC_Back_Center_TWNG_comp} show a complete set of data, averaged over two replicates, for the instrumented, non combustible wall natural gas burner experiments.

\begin{figure}[ht!]
	\centering
	\includegraphics[width=2.5in]{../Figures/TWNG01_TC_Plume_Avg}
	\includegraphics[width=2.5in]{../Figures/TWNG03_TC_Plume_Avg}\\
	\includegraphics[width=2.5in]{../Figures/TWNG05_TC_Plume_Avg}
	\includegraphics[width=2.5in]{../Figures/TWNG07_TC_Plume_Avg}\\
	\caption[Plume temperatures along the centerline of the natural gas fueled burner]{Plume temperatures along the centerline of the natural gas fueled burner. The temperatures are averages of two replicate experiments. Each graph displays data from a different distance between the edge of the burner and the instrumented wall.  Beginning with the top graph position and moving down the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m respectively.}
	\label{TC_Plume_TWNG_comp}
\end{figure}

\begin{figure}[ht!]
	\centering
		\includegraphics[width=2.5in]{../Figures/TWNG01_TC_Surface_Center_Avg}
		\includegraphics[width=2.5in]{../Figures/TWNG03_TC_Surface_Center_Avg}\\
		\includegraphics[width=2.5in]{../Figures/TWNG05_TC_Surface_Center_Avg}
		\includegraphics[width=2.5in]{../Figures/TWNG07_TC_Surface_Center_Avg}\\
	\caption[Wall Temperatures along the centerline of the natural gas fueled burner]{Wall Temperatures along the centerline of the natural gas fueled burner. The temperatures are averages of two replicate experiments. Each graph displays data from a different distance between the edge of the burner and the instrumented wall.  Beginning with the top graph position and moving down the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m, respectively.}
	\label{TC_Surf_Cent_TWNG_comp}
\end{figure}

\begin{figure}[ht!]
	\centering
	\includegraphics[width=2.5in]{../Figures/TWNG01_TC_Surface_Offset_Avg}
	\includegraphics[width=2.5in]{../Figures/TWNG03_TC_Surface_Offset_Avg}\\
	\includegraphics[width=2.5in]{../Figures/TWNG05_TC_Surface_Offset_Avg}
	\includegraphics[width=2.5in]{../Figures/TWNG07_TC_Surface_Offset_Avg}\\
	\caption[Wall Temperatures along the edge of the natural gas fueled burner]{Wall Temperatures along the right edge of the natural gas fueled burner, when facing the wall. The temperatures are averages of two replicate experiments. Each graph displays data from a different distance between the edge of the burner and the instrumented wall.  Beginning with the top graph position and moving down the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m, respectively.}
	\label{TC_Surf_Edge_TWNG_comp}
\end{figure}

\begin{figure}[ht!]
	\centering
	\includegraphics[width=2.5in]{../Figures/TWNG01_TC_Back_Center_Avg}
	\includegraphics[width=2.5in]{../Figures/TWNG03_TC_Back_Center_Avg}\\
	\includegraphics[width=2.5in]{../Figures/TWNG05_TC_Back_Center_Avg}
	\includegraphics[width=2.5in]{../Figures/TWNG07_TC_Back_Center_Avg}\\
	\caption[Backside wall Temperatures along the centerline of the natural gas fueled burner]{Backside wall Temperatures along the centerline of the natural gas fueled burner. The temperatures are averages of two replicate experiments. Each graph displays data from a different distance between the edge of the burner and the instrumented wall.  Beginning with the top graph position and moving down the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m, respectively.}
	\label{TC_Back_Center_TWNG_comp}
\end{figure}

@@

Figure~\ref{NCTW_Surf_Center_comp} shows the temperatures of the array of thermocouples located along the vertical centerline of the burners for each of the three fuels. Reviewing the graph for each fuel demonstrates how the temperature at the wall decreased as the position of the burner was moved away from the wall. The figures  The beads of the thermocouples were installed in the wall and were flush with the wall surface. In each graph, the average peak temperature of the replicate experiments, for each thermocouple position is shown. Moving from 0.2~m above the top surface of the burner up to 2.2~m above the top of the burner surface in increments of 0.2~m. As with the data analysis for the source fires, the peak averages were based on a 30~s period bounding the peak for the natural gas and gasoline fires.  A 15~s interval bounding the peak was used for the polyurethane, in order to be more representative of the peak conditions.

@@

@@

\section{Experimental Arrangement}
This series of experiments combined the instrumentation used in the non-combustible wall experiments with gypsum wallboard, which has a combustible paper surface.  A free standing, steel frame with a panel of gypsum wallboard attached to the surface was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was moved toward the wall until the burner was against the wall.  If the results indicated that no burn pattern would result from a given combination of fuel and position, then that fuel and position combination were eliminated from the remainder of the test matrix.

@@

According to the United States Geological Survey, most gypsum in the United States is used to make wallboards for homes and other structures.  A typical home in the United States contains more than 7 metric tons of gypsum ref USGS website http://minerals.usgs.gov/minerals/pubs/commodity/gypsum/.  As defined by ASTM, gypsum wallboard is gypsum board designed for use on walls, ceilings, or partitions and that affords a surface suitable to recieve decoration  ref ATSM 1396. The ASTM standard further describes gypsum wallboard as having a noncombustible core, surfaced with paper bonded to the core. By weight, wallboard is composed of more than 85~\% gypsum, less than 10~\% cellulose, less than 3~\% starch and approximately 1~\% of both crystalline silica and fibrous glass.  Chemically gypsum can be described as calcium sulfate dihydrate, (CaSO$_{4}$2H$_{2}$O) .  By weight pure gypsum is approximately 79~\% calcium sulfate and 21~\% chemically bound water ref Lawson 1977.

The gypsum wallboard used in these experiments is referred to as ``regular'' as opposed to a gypsum wallboard `Type C' or `Type X' which are used in fire rated assemblies.   The gypsum wallboard used in these experiments was purchased from same manufacturer in an effort to have the most similar panels for use between the experiments.  The physical features of the gypsum wallboard panels are given in the table below.

@@

After receipt of the gypsum wallboard panels, the panels were laid out on the floor of the NIST Large Fire Research Laboratory, finished paper side up and painted with a roller.  The first coat was a latex primer.  Once the primer dried a top coat of white latex paint was applied.  After the panels dried paint thickness measurements were made with a ultra sonic XXXX.  Some of the panels were then cut-up into samples think about this perhaps this goes into the FP Comp

>>>>

\Title{Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard}
\Author{Daniel Madrzykowski}
\Year{2015}

@@

A free standing steel frame wall with a 12.7~mm thick, 2.44~m high by 1.22~m wide, calcium silicate board face was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux. Marinite Type I is the commercial name of the non-combustible board used in these experiments.  It is normally used as a fire-resistant thermal insulation that also has structural strength. 90~\% of the board is composed of calcium silicate and calcium meta-silicate.  The remaining 10~\% of the board is composed of a combination of organic fibers, fiber glass filament and crystalline silica (quartz)~\cite{BNZ:2012}.  Material properties that impact heat transfer were selected from the manufacturer's literature and are presented in the table below~\cite{Marinite:1997}.

\begin{table}
	\centering
\begin{tabular}{|c|c|c|c|c|}
	\hline Density & \multicolumn{2}{c|}{Thermal Conductivity} & \multicolumn{2}{c|}{Specific Heat}   \\
    \hline (kg/m$^3$) & Temperature ($^\circ$C) & (W/mK)  & Temperature ($^\circ$C)  & (kJ/kg $^\circ$C) \\
	\hline 737  & 24 & 0.13 & 93 & 0.28 \\
	\hline  & 205 & 0.12  & 205 & 0.30 \\
	\hline  & 316 & 0.11 & 316 & 0.32 \\
	\hline  & 425 & 0.12 & 425 & 0.34 \\
	\hline  & 538 & 0.12 &  &  \\
	\hline
\end{tabular}
\caption[Thermal properties for Marinite I]{Thermal properties for Marinite I, as reported by the manufacturer BNZ~\cite{Marinite:1997}
\label{tab:Marinite_Thermal_Properties}
\end{table}

@@

Figures~\ref{TC_Plume_TWNG_comp} through \ref{TC_Back_Center_TWNG_comp} show a complete set of data, averaged over two replicates, for the instrumented, non combustible wall natural gas burner experiments.  Comparing the temperatures of the plume at the most distant point from the wall, 0.61~m and with the plume temperatures with the burner contacting the wall in Figure~\ref{TC_Plume_TWNG_comp} shows an increase.  However for the burner position, 0.15~m from the wall, a decrease in plume temperatures occurs after approximately 3 minutes after ignition.  A review of the heat release rates and the test videos do not yield an clues as to the reason for this decrease which occurred in both experiments.  In both experiments the heat release rate was steady and the video did not show any change in flame movement.  
Figures~\ref{TC_Surf_Cent_TWNG_comp} and \ref{TC_Surf_Edge_TWNG_comp} show consistent trends of the thermocouple temperatures in the vertical arrays on the wall increasing as the burner was moved toward the wall.  Figure~\ref{TC_Back_Center_TWNG_comp} shows that the peak temperatures aligned with the vertical center of the burner on the backside of the gypsum wallboard peaked at approximately 100~$^\circ$C.

\begin{figure}[ht!]
	\centering
	\includegraphics[width=2.5in]{../Figures/TWNG01_TC_Plume_Avg}
	\includegraphics[width=2.5in]{../Figures/TWNG03_TC_Plume_Avg}\\
	\includegraphics[width=2.5in]{../Figures/TWNG05_TC_Plume_Avg}
	\includegraphics[width=2.5in]{../Figures/TWNG07_TC_Plume_Avg}\\
	\caption[Plume temperatures along the centerline of the natural gas fueled burner]{Plume temperatures along the centerline of the natural gas fueled burner. The temperatures are averages of two replicate experiments. Each graph displays data from a different distance between the edge of the burner and the instrumented wall.  Beginning with the top left graph position and viewing left to right the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m respectively.}
	\label{TC_Plume_TWNG_comp}
\end{figure}

\begin{figure}[ht!]
	\centering
		\includegraphics[width=2.5in]{../Figures/TWNG01_TC_Surface_Center_Avg}
		\includegraphics[width=2.5in]{../Figures/TWNG03_TC_Surface_Center_Avg}\\
		\includegraphics[width=2.5in]{../Figures/TWNG05_TC_Surface_Center_Avg}
		\includegraphics[width=2.5in]{../Figures/TWNG07_TC_Surface_Center_Avg}\\
	\caption[Wall Temperatures along the centerline of the natural gas fueled burner]{Wall Temperatures along the centerline of the natural gas fueled burner. The temperatures are averages of two replicate experiments. Each graph displays data from a different distance between the edge of the burner and the instrumented wall.  Beginning with the top left graph position and viewing left to right the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m respectively.}
	\label{TC_Surf_Cent_TWNG_comp}
\end{figure}

\begin{figure}[ht!]
	\centering
	\includegraphics[width=2.5in]{../Figures/TWNG01_TC_Surface_Offset_Avg}
	\includegraphics[width=2.5in]{../Figures/TWNG03_TC_Surface_Offset_Avg}\\
	\includegraphics[width=2.5in]{../Figures/TWNG05_TC_Surface_Offset_Avg}
	\includegraphics[width=2.5in]{../Figures/TWNG07_TC_Surface_Offset_Avg}\\
	\caption[Wall Temperatures along the edge of the natural gas fueled burner]{Wall Temperatures along the right edge of the natural gas fueled burner, when facing the wall. The temperatures are averages of two replicate experiments. Each graph displays data from a different distance between the edge of the burner and the instrumented wall.  Beginning with the top left graph position and viewing left to right the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m respectively.}
	\label{TC_Surf_Edge_TWNG_comp}
\end{figure}

\begin{figure}[ht!]
	\centering
	\includegraphics[width=2.5in]{../Figures/TWNG01_TC_Back_Center_Avg}
	\includegraphics[width=2.5in]{../Figures/TWNG03_TC_Back_Center_Avg}\\
	\includegraphics[width=2.5in]{../Figures/TWNG05_TC_Back_Center_Avg}
	\includegraphics[width=2.5in]{../Figures/TWNG07_TC_Back_Center_Avg}\\
	\caption[Backside wall Temperatures along the centerline of the natural gas fueled burner]{Backside wall Temperatures along the centerline of the natural gas fueled burner. The temperatures are averages of two replicate experiments. Each graph displays data from a different distance between the edge of the burner and the instrumented wall. Beginning with the top left graph position and viewing left to right the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m respectively.}
	\label{TC_Back_Center_TWNG_comp}
\end{figure}

Figures~\ref{HF_Center_TWNG_comp} and \ref{HF_Edge_TWNG_comp} provide the heat flux measurements for the two vertical arrays of gauges positioned along the centerline and the edge of the natural gas burner.  The series of graphs show that the peak heat flux increases by as the burner is moved toward the wall. One exception is when the burner was moved from the 0.15~m position to the 0.0~m from the wall. In this comparison the peak heat flux values for the gauges located above the edge of the burner do not uniformly increase.  In fact the peak heat flux actually decreased as the burner was moved from the 0.15~m position to the 0.0~m position. This decrease was due to the change in view angle between the flame closest to the burner and the lower heat flux gauges at the edge of the burner.

\begin{figure}[ht!]
	\centering
	\includegraphics[width=2.5in]{../Figures/HF_Center_TWNG_0.61m_Reps_Avg}
	\includegraphics[width=2.5in]{../Figures/HF_Center_TWNG_0.31m_Reps_Avg}\\
	\includegraphics[width=2.5in]{../Figures/HF_Center_TWNG_0.15m_Reps_Avg}
	\includegraphics[width=2.5in]{../Figures/HF_Center_TWNG_0m_Reps_Avg}\\
	\caption[Heat flux at the wall along the centerline of the natural gas fueled burner]{Heat flux at the wall along the centerline of the natural gas fueled burner. The heat fluxes are averages of two replicate experiments. Each graph displays data from a different distance between the edge of the burner and the instrumented wall. Beginning with the top left graph position and viewing left to right the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m respectively.}
	\label{HF_Center_TWNG_comp}
\end{figure}

\begin{figure}[ht!]
	\centering
	\includegraphics[width=2.5in]{../Figures/HF_Offset_TWNG_0.61m_Reps_Avg}
	\includegraphics[width=2.5in]{../Figures/HF_Offset_TWNG_0.31m_Reps_Avg}\\
	\includegraphics[width=2.5in]{../Figures/HF_Offset_TWNG_0.15m_Reps_Avg}
	\includegraphics[width=2.5in]{../Figures/HF_Offset_TWNG_0m_Reps_Avg}\\
	\caption[Heat flux at the wall along the edge of the natural gas fueled burner]{Heat flux at the wall along the edge of the natural gas fueled burner. The heat fluxes are averages of two replicate experiments. Each graph displays data from a different distance between the edge of the burner and the instrumented wall. Beginning with the top left graph position and viewing left to right the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m respectively.}
	\label{HF_Edge_TWNG_comp}
\end{figure}

In the following sections, the data is compiled so that averaged peak values, of a given measure, for each position from the wall will be presented on a single graph for a given fuel type.

@@

Figure~\ref{NCTW_Surf_Center_comp} shows the temperatures of the array of thermocouples located along the vertical centerline of the burners for each of the three fuels. Reviewing the graph for each fuel demonstrates how the temperature at the wall decreased as the position of the burner was moved away from the wall. The beads of the thermocouples were installed in the wall and were flush with the wall surface. In each graph, the average peak temperature of the replicate experiments, for each thermocouple position is shown. Moving from 0.2~m above the top surface of the burner up to 2.2~m above the top of the burner surface in increments of 0.2~m. As with the data analysis for the source fires, the peak averages were based on a 30~s period bounding the peak for the natural gas and gasoline fires.  A 15~s interval bounding the peak was used for the polyurethane, in order to be more representative of the peak conditions.

@@

\Section{Summary}

These instrumented non-combustible wall experiments provide a data set for the fires conducted that can be compared against ignition data for gypsum wallboard from the literature to determine the potential for ignition of the paper surface of gypsum wallboard at a given distance from the wall with a given fuel. The Ignition Handbook by Babrauskas is the most comprehensive set of ignition data and related information ever written~\cite{Babrauskas:2003}.  Table 19 in the Ignition Handbook includes calculated ignition temperatures from experimental data from two separate source. Quintiere and Harkleroad conducted a study with the Lateral Ignition and Flame Spread (LIFT) apparatus, which they developed ~\cite{ASTM_E1321},\cite{Quitiere:1984}.  
The ignition temperature value calculated from the LIFT data for regular gypsum wallboard, 12.7~mm thick was 565 $\^circ$C with a critical heat flux for ignition of 35~kW/m^2.

The second source, was a series of cone calorimeter experiments conducted by Dillion~\cite{ASTM_E1354},~\cite{Dillion:1998}. The ignition temperature calculated from the cone calorimeter experiments for regular gypsum wallboard, 12.5~mm thick was 515 $\^circ$C with a minimum heat flux required for ignition of 26~kW/m^2.

A review of the graphs of the peak average temperatures at the wall, burner centerline array, shows that all of the fuels had at least one thermocouple position with a peak of approximately 500 $\^circ$C when the burner was against the wall.  At 0.15~m away from the wall the peak average temperature from this data set was provided by the gasoline fueled fire and it was approximately 350 $\^circ$C.  Based on temperature alone, the fires examined would only have the potential to ignite the paper surface of the gypsum wallboard when the burner was in contact with the wall.  The peak average temperatures at the wall, burner edge array, would indicate that the paper may not ignite at the burner edge.  The peak average temperature of approximately 400 $\^circ$C was generated by the gasoline fueled fire when the burner was positioned against the wall.

Of course the temperatures of the thermocouples in the wall as a result the of the energy transferred to the wall and the thermo-physical properties of the wall material. Assuming that the properties of calcium silicate board and the gypsum board are similar, let's review the heat flux graphs. All of the fuels generated heat fluxes in excess of 35~kW/m^2 at the wall, burner centerline array when the burners were against the wall.  The gasoline and the polyurethane fueled burners also exposed the wall to heat fluxes higher than 35~kW/^2 at 0.15~m from the wall.  However only the burner against the wall positions provided a combination of heat flux and temperature that had the potential to ignite the paper on the gypsum wallboard based on the handbook data.

@@

\section{Experimental Arrangement}
This series of experiments combined the instrumentation used in the non-combustible wall experiments with gypsum wallboard, which has a combustible paper surface. These experiments enable us to examine our ignition/no ignition predictions from the previous experiments.  A free standing, steel frame with a panel of gypsum wallboard attached to the surface was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was moved toward the wall until the burner was against the wall.  If the results indicated that no burn pattern would result from a given combination of fuel and position, then that fuel and position combination were eliminated from the remainder of the experiments in the study.

@@

According to the United States Geological Survey, most gypsum in the United States is used to make wallboards for homes and other structures.  A typical home in the United States contains more than 7 metric tons of gypsum ref USGS website http://minerals.usgs.gov/minerals/pubs/commodity/gypsum/.  As defined by ASTM, gypsum wallboard is gypsum board designed for use on walls, ceilings, or partitions and that affords a surface suitable to receive decoration  ref ATSM 1396. The ASTM standard further describes gypsum wallboard as having a noncombustible core, surfaced with paper bonded to the core. By weight, wallboard is composed of more than 85~\% gypsum, less than 10~\% cellulose, less than 3~\% starch and approximately 1~\% of both crystalline silica and fibrous glass.  Chemically gypsum can be described as calcium sulfate dihydrate, (CaSO$_{4}$2H$_{2}$O) .  By weight pure gypsum is approximately 79~\% calcium sulfate and 21~\% chemically bound water ref Lawson 1977.

The gypsum wallboard used in these experiments is referred to as ``regular'' as opposed to a gypsum wallboard `Type C' or `Type X' which are used in fire rated assemblies.   The gypsum wallboard used in these experiments was purchased from same manufacturer in an effort to have the most similar panels for use between the experiments. Each 1.22~m by 2.44~m by 12.7~mm thick panel weighed approximately 30.5~kg.  The physical features of the gypsum wallboard panels are given in the table below.

@@

After receipt of the gypsum wallboard panels, the panels were laid out on the floor of the NIST Large Fire Research Laboratory, finished paper side up and painted with a roller.  The first coat was a latex primer, http://buyat.ppg.com/rep_pafpainttools_files/Pghpaints/tdb/78200.pdf.  Once the primer dried a top coat of white latex paint was applied http://buyat.ppg.com/rep_pafpainttools_files/Olympic/tdb/78010.pdf .  After the panels dried, they were weighed. The average mass of paint on the finished surface of the gypsum wallboard was 0.33~kg. Then the paint thickness was estimated an ultra sonic XXXX cite - in office, the thickness of the paint varied from 0.05~mm to 0.13~mm.

Some of the panels were then cut-up into samples think about this perhaps this goes into the FP Comp
